---
title: "Project 1"
author: "Caroline Hutchings"
date: today
format: html
editor: visual
---

## Census Data API

The purpose of this project is to create a reusable set of functions that interact with the United States Census Bureau API. Rather than manually downloading data for a specific year, variable, or geographic area, these functions allow users to dynamically request data from the Census Public Use Microdata Sample (PUMS) dataset. The project focuses on four survey years, 2021 through 2024, and demonstrates how metadata can be used to automate variable selection, labeling, and data cleaning.

To accomplish this goal, I first developed a function to retrieve metadata describing the available variables. Next, I created helper functions to convert the raw API responses into usable data structures and to properly interpret coded values. Finally, I combined these components into a larger function capable of requesting, cleaning, and returning Census data in a user-friendly format.

The first step in accomplishing the task is to load libraries that will potentially be used throughout the document. The `httr` package is used to communicate with the Census API, while `jsonlite` parses JSON responses into R objects. The `dplyr` and `tidyverse` packages provide data wrangling functionality, and `lubridate` is used to work with time variables. Together, these packages make it possible to collect, clean, and transform the Census data efficiently.

```{r}
library(httr)
library(jsonlite)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
```

# Part 1: Functions

Before requesting data from the API, it is important to understand the available variables and their associated metadata. The Census API provides a file containing descriptions, labels, and possible values for each variable. Since some variable names differ across survey years, I created a function that automatically retrieves the correct metadata and selects only the variables relevant to this project.

```{r}
#Create a function to get all the variables named funcation variable list (fun_var_list)
fun_var_list <- function(year) {
  
  # Exclude any years outside of the following    
  if (!year %in% c(2021, 2022, 2023, 2024)) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # Use the API to get the data, use paste to add the year as a variable determined by the function    
  response <- GET(
    paste0("https://api.census.gov/data/",
           year,
           "/acs/acs1/pums/variables.json"))
  
  # Parse the data from JSON  
  parse_var <- fromJSON(rawToChar(response$content))
  
  # Pull out the variables required.  Choose the names depending on the year requested because they change for certain years.  
  
  if (year %in% c(2021, 2022)) {
    required_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH","SCHL", "SEX", "ST", "REGION", "DIVISION")
    
  } else {
    required_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP","JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")
  }
  
  #create a list that includes only the required variables.    
  var_list <- parse_var$variables[required_vars]
  
  #return the list  
  return(var_list)
}
```

### Test the function

The function was tested using each of the four survey years included in the project. For each year, the function successfully retrieved the metadata from the Census API and returned a list containing the selected variables and their associated information. This confirms that the function can dynamically build the correct API request and extract the desired variable information regardless of which of the supported years is selected.

An important part of testing was verifying that the function properly handles invalid user input. To accomplish this, I attempted to run the function using the year 2025. Since the project only supports the years 2021 through 2024, the function correctly stopped execution and returned an error message. This error handling prevents users from requesting years for which the function was not designed and helps ensure that only valid API requests are submitted.

```{r}
#Run the function for each year and create a list
var_21 <- fun_var_list(2021)
var_22 <- fun_var_list(2022)
var_23 <- fun_var_list(2023)
var_24 <- fun_var_list(2024)
#Try a year out of range, should produce an error.
#var_25 <- fun_var_list(2025)
```

### View the List

After creating the variable lists, I used the `View()` function to explore their structure and verify that the metadata was successfully retrieved. Examining the list revealed that each variable contains several pieces of information, including a descriptive label and, for categorical variables, a collection of possible coded values stored in the `values$item` section. By inspecting the metadata, I was able to identify where the variable descriptions and response categories were stored. For example, the `SEX` variable contains two possible values, Male and Female, making it straightforward to convert the coded responses into readable labels. In contrast, the `JWAP` variable, which represents time of arrival at work, contains 286 possible values in its `items` list. Each of these values corresponds to a specific arrival time interval. The large number of categories for variables such as `JWAP` highlights the importance of using metadata to correctly interpret and label the Census data.

Understanding the structure of these metadata objects was an important step because many Census variables are stored as numeric codes rather than meaningful labels. The information contained in the metadata is later used by the helper functions to convert coded values into more interpretable formats.

```{r}
View(var_21)
```

## Combine the Years into a List

Since the metadata structure is similar across years, it is convenient to store all four years in a single object. Combining the metadata into one list makes it easier to reference the correct year later in the project. Saving the object as an RDS file also prevents unnecessary API requests because the metadata can be loaded directly from disk. The RDS file was a new file type to me and I had to do some research on how to create the RDS file.

```{r}
#Combine all the lists into one list
all_var_list <- list(variables_21 = var_21, 
                     variables_22 = var_22, 
                     variables_23 = var_23, 
                     variables_24 = var_24)

#Save the list as an RDS file in the data folder
saveRDS(all_var_list, "data/variable_list.RDS")
```

## Use API 

Once the metadata was available, the next step was learning how the Census API returns actual survey data. I began by creating a single test request and examining the structure of the response. Understanding the response format was necessary before building a general-purpose function to automate data collection.

```{r}
#Using the api get the data and parse it from JSON

get_API_2021 <- GET("https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,JWAP,MAR&for=division:*&SCHL=24&key=96f181e4ada2cf8ebea839ffd9fa684508d8fc09")

#Identify the names in order to find where the data is stored.
names(get_API_2021)

parsed_census <- fromJSON(rawToChar(get_API_2021$content))
```

Because every Census API request returns data in a similar format, I created a helper function to perform the conversion process automatically. Encapsulating this logic inside a function reduces repeated code and simplifies future data requests.

## Helper Function to Parse Data

The raw API response is returned as JSON, which is not a user friendly format for statistical analysis. Once the JSON response was parsed, I explored the organization of the returned dataset and identified that the API stores column names in the first row rather than as traditional variable names. The remaining rows contained the actual observations. Recognizing this pattern allowed me to develop a helper function that systematically extracts the column names and converts the data into a clean tibble ready for analysis.

```{r}
#Helper function to create a tibble from parsed data
helper <- function(parsed_data){
  new_tibble <- as_tibble(parsed_data[-1, ], 
                          .name_repair='minimal')
  names(new_tibble) <- parsed_data[1, ]
  return(new_tibble)
}
#Use helper function to create a tibble for the parsed census data
helper(parsed_census)
```

Several Census variables record times using coded intervals rather than exact clock times. For example, a code may represent an arrival time between two specified hours. To make these variables easier to analyze, I created a helper function that converts each interval into its midpoint. This provides a single representative time while preserving the overall timing information contained in the original data.

## Helper Function to find Midpoint Time 

Create the time helper function that finds the midpoint of the time in the given time interval. There are several different variables available to test the time helper function on. One of the variables JWMNP is essentially a null value variable as the only value is they did not travel to work that day. I removed the day portion of the time/date variable since this information was not provided and only returned the hour:minute:second portion of the date/time value.

```{r}
#loading the all variables object so the items can be used to identify the labels for the variables.
all_var_list <- readRDS("data/variable_list.RDS")


#Create the helper time function
helper_time <- function(code, var_name, year_metadata){
  #code 0 as NA based on the item for the value
  if(is.na(code) || code == "0"){
    return(NA)
  }
  #Pull the label from the variable using its value
  label <-
    year_metadata[[var_name]]$values$item[[as.character(code)]]
  
  if(is.null(label)){
    return(NA)
  }
  #Format the time into a usable format by removing to and fixing p.m.
  times <- strsplit(label, " to ")[[1]]
  times <- gsub("\\.","", times)
  
  start <- parse_date_time(times[1], "I:M p")
  end   <- parse_date_time(times[2], "I:M p")
  
  #Finding the midpoint of the two times in the label
  midpoint <- start + (end - start)/2
  
  #Return the midpoint in hour and minute format
  return(hm(format(midpoint, "%H:%M")))
}


#Using the helper time function to make sure it works and returns the time in hour:minute format
helper_time("198","JWAP" , all_var_list$variables_21)
```

## Helper Function for Categorical Variables

Many Census variables are stored as coded categories rather than descriptive labels. Before creating the helper function, I reviewed the metadata and confirmed that variables such as FER, HHL, SCH, SCHL, and SEX are categorical and should be stored as factors, they don't have too many levels that factors would be meaningless. The helper function uses the metadata to replace coded values with readable labels and convert the variable into a factor.

```{r}
#Checking validity of creating factors for the variables
length(all_var_list$variables_24$FER$values$item)
length(all_var_list$variables_24$HHL$values$item)
length(all_var_list$variables_24$SCH$values$item)
length(all_var_list$variables_24$SCHL$values$item)
all_var_list$variables_24$SCHL$values$item
```

Write a helper function to force the categorical variables into factors.

```{r}
# Create a factor helper function
helper_factor <- function(x, labels){
  factor(
    x,
    levels = names(labels),
    labels = unlist(labels)
  )
}
```

## Function to Use API 

After creating the helper functions, I combined them into a single function that retrieves Census data and returns a cleaned tibble. The function includes default values, validates user inputs, and automatically selects the correct metadata for the requested year.

Because the state variable is named `ST` in 2021 and 2022 but `STATE` in 2023 and 2024, the function adjusts automatically based on the survey year. After downloading the data, it converts numerical variables, applies midpoint calculations to time variables, and converts categorical variables into labeled factors.

```{r}
api_key <- "96f181e4ada2cf8ebea839ffd9fa684508d8fc09"


#create a function to get the data from the api and create a tibble.
get_api_data <- function(
    #default values
  year = 2024,
  num_var = "AGEP",
  cat_var = "SEX",
  geography = "State",
  geo_value = NULL){
  
  # Check year is in the given 4 years
  if(length(year) != 1 ||
     !year %in% c(2021, 2022, 2023, 2024)){
    stop("Year must be 2021, 2022, 2023, or 2024")
  }
  
  # Select metadata year
  if(year == 2021){
    year_metadata <- all_var_list$variables_21
  } else if(year == 2022){
    year_metadata <- all_var_list$variables_22
  } else if(year == 2023){
    year_metadata <- all_var_list$variables_23
  } else {
    year_metadata <- all_var_list$variables_24
  }
  
  # Check numeric variable is one of the set variables
  allowed_num <- c(
    "AGEP",
    "GASP",
    "GRPIP",
    "JWAP",
    "JWDP",
    "JWMNP"
  )
  
  if(length(num_var) != 1){
    stop("Specify exactly one numeric variable")
  }
  
  if(!num_var %in% allowed_num){
    stop("Invalid numeric variable")
  }
  
  # PWGTP always included
  api_vars <- c(num_var, "PWGTP")
  
  # Check categorical variable is from given list
  allowed_cat <- c(
    "FER",
    "HHL",
    "SCH",
    "SCHL",
    "SEX"
  )
  
  if(length(cat_var) != 1){
    stop("Specify exactly one categorical variable")
  }
  
  if(!cat_var %in% allowed_cat){
    stop("Invalid categorical variable")
  }
  #define the available variables to choose from
  api_vars <- c(api_vars, cat_var)
  
  # Check geography is one of the values specified
  allowed_geo <- c(
    "Region",
    "Division",
    "State"
  )
  
  if(length(geography) != 1){
    stop("Specify exactly one geography level")
  }
  
  if(!geography %in% allowed_geo){
    stop("Geography must be Region, Division, or State")
  }
  
  # Default geography value
  if(is.null(geo_value)){
    
    geo_defaults <- c(
      State = "06",
      Division = "9",
      Region = "4"
    )
    
    geo_value <- unname(
      geo_defaults[geography]
    )
    
  }
  
  # Validate geography value
  if(geography == "State"){
    
    if(year %in% c(2021, 2022)){
      geo_var <- "ST"
    } else {
      geo_var <- "STATE"
    }
    
  } else {
    
    geo_var <- toupper(geography)
    
  }
  
  
  valid_geo <- names(
    year_metadata[[geo_var]]$values$item
  )
  
  if(tolower(geo_value) != "all" &&
     !geo_value %in% valid_geo){
    stop("Invalid geography value")
  }
  
  # Build URL
  if(tolower(geo_value) == "all"){
    
    url <- paste0(
      "https://api.census.gov/data/",
      year,
      "/acs/acs1/pums?get=",
      paste(api_vars, collapse = ","),
      "&for=",
      tolower(geography),
      ":*",
      "&key=",
      api_key
    )
    
  } else {
    
    url <- paste0(
      "https://api.census.gov/data/",
      year,
      "/acs/acs1/pums?get=",
      paste(api_vars, collapse = ","),
      "&for=",
      tolower(geography),
      ":",
      geo_value,
      "&key=",
      api_key
    )
    
  }
  
  response <- GET(url)
  
  parsed <- fromJSON(
    rawToChar(response$content)
  )
  
  census_tbl <- helper(parsed)
  
  # Convert numeric/time variable
  if(num_var %in% c(
    "JWAP",
    "JWDP",
    "JWMNP")){
    
    census_tbl[[num_var]] <- sapply(
      census_tbl[[num_var]],
      helper_time,
      var_name = num_var,
      year_metadata = year_metadata
    )
    
  } else {
    
    census_tbl[[num_var]] <-
      as.numeric(census_tbl[[num_var]])
    
  }
  
  # Convert PWGTP
  census_tbl$PWGTP <-
    as.numeric(census_tbl$PWGTP)
  
  # Convert categorical variable
  labels <- year_metadata[[cat_var]]$values$item
  
  census_tbl[[cat_var]] <-
    helper_factor(
      census_tbl[[cat_var]],
      labels
    )
  
  # Add census class
  class(census_tbl) <-
    c("census", class(census_tbl))
  
  return(census_tbl)
  
}

```

### Test the API Function

I tested the function using several combinations of arguments, including different numerical variables, categorical variables, and time variables. These tests confirmed that the function correctly retrieves the requested data and applies the appropriate conversions before returning a tibble.

```{r}
get_api_data()

# Different categorical variable
get_api_data(cat_var = "SCHL")

get_api_data(num_var = "GRPIP")

get_api_data(num_var="JWAP")
```

## Create a Function for Multiple Years

To allow data from multiple years to be analyzed together, I created a function that repeatedly calls the single-year function and combines the results. A year variable is added to each dataset before the data are merged, making it possible to identify the source year of each observation.

```{r}

# Create a function that combines more than 1 year
get_api_data_multi <- function(
    years,
    num_var = "AGEP",
    cat_var = "SEX",
    geography = "State",
    geo_value = NULL){
  
  #set table to empty
  final_tbl <- NULL
  #create a loop to go through for each given year.
  for(year in years){
    
    temp <- get_api_data(
      year = year,
      num_var = num_var,
      cat_var = cat_var,
      geography = geography,
      geo_value = geo_value
    )
    
    temp$year <- year
    
    #first pass fill the table with the temp each additional pass append the new data to the table.
    if(is.null(final_tbl)){
      final_tbl <- temp
    } else {
      final_tbl <- bind_rows(final_tbl, temp)
    }
    
  }
  
  return(final_tbl)
  
}
```

### Test the Function

The multi-year function was tested using data from 2021 and 2022. The results showed that data from both years were successfully retrieved, combined, and returned as a single tibble. This verified that the functions work together to automate the collection and preparation of Census data across multiple years.

```{r}
get_api_data_multi(years = c(2021, 2022))
```
